TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2019-03-13 paper/draft.tex
modify wording

Why is there a problem determining M dwarf metallicities (and especially for high res \apogee\ spectra)? Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled.

In particular, parameters such as metallicity can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths. A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.

@@

The data-driven approach of \thecannon\ is ideal in certain cases:
if stellar labels are only known for a small number of stars
but there are spectra taken for many more;
if it is computationally expensive to obtain labels for a star, but
there are many stars that need labels;
if there are spectral models that work in one wavelength range or resolution but not in another.
The context of this paper is the last:
There are working and reliable models of M dwarf stars in the optical, which have been
used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.
\thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.

@@

A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
\begin{enumerate}
\item \citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs
\item \citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances
\item \citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs
\item \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
\item \citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries
\end{enumerate}

>>>>

Why is there a problem determining M dwarf metallicities (and especially for high res \apogee\ spectra)? Emphasize issues: (1) models dont accurately reproduce data; (2) empirical standards are relative not absolute, and they are sparesely sampled. 
Methods for determining metallicity of M dwarfs: (1) from atmospheric models of M dwarfs, (2) empirical methods

In particular, the metallicity abundance can be difficult to determine in M dwarfs, as overlapping lines in a spectrum and lack of continuum make it difficult to define features and extract their line strengths. A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically similar pairs are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
However these metallicity callibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant callibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.

@@

The data-driven approach of \thecannon\ is ideal in certain cases:
if stellar labels are only known for a small number of stars
but there are spectra taken for many more;
if it is computationally expensive to obtain labels for a star, but
there are many stars that need labels;
if there are spectral models that work in one wavelength range or resolution but not in another.
The context of this paper is the last:
there are precise callibrations for metallicities of M dwarf stars in the optical, which have been
used successfully, but (as we note above) there isn't as much success with such modeling in the infrared, and not at \apogee\ resolution.
\thecannon\ bridges this gap: it doesn't require using specific lines; instead it allows us to determine labels from a lot of low level metallicity information present in thousands of lines, and as we demonstrate, does so with very good precision.

@@

A number of M dwarf studies have already been conducted in \apogee, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements. \citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs. \citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances. \citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs. \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity). \citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries.
